---
title: "Fig 3 - COVID-19 signature validation"
output: rmarkdown::github_document
---

Here we assess the generalization performance (including both detection and cross-reactivity) of the identified 11-gene COVID-19 signature in independent studies. We then compare the resulting performance with four previously published signatures.

### Signature performance in GSE163151
This dataset includes samples from COVID-19 patients, from other viral 
and bacterial infections, and from healthy controls. Thus, it makes it possible to simultaneously test  COVID-19 detection and cross-reactivity. 
```{r, message = FALSE}
library(dplyr)
library(ggplot2)
source("../scripts/helper_functions.R")

# signature definition
COVID19_signature_up <-
  c("PIF1", "GUCD1", "EHD3", "TCEAL3", "BANF1")
COVID19_signature_down <-
  c("ARAP2", "SLC25A46", "SLK", "ROCK2", "TVP23B", "DOCK5")

# reading dataset
dataset <-
  readRDS(file = "../../data/mRNA_studies/GSE163151_preprocessed.RDS")
COVID19_signature_up_expression <- dataset$X[COVID19_signature_up, ]
COVID19_signature_down_expression <-
  dataset$X[COVID19_signature_down, ]

# computing signature scores for each sample
sample_score <- scale(
  apply(COVID19_signature_up_expression, 2, function(x) {
    geom_mean(x)
  }) -
    apply(COVID19_signature_down_expression, 2, function(x) {
      geom_mean(x)
    })
)

dataset$y[, "disease state:ch1"] <- factor(
  dataset$y[, "disease state:ch1"],
  levels = c(
    "Donor control",
    "COVID-19",
    "Viral acute respiratory illness",
    "Bacterial sepsis"
  )
)

score_df <- data.frame(
  score = sample_score,
  group = dataset$y[, "disease state:ch1"]
) %>%
  dplyr::mutate(group = relevel(group, ref = "Donor control"))


# ROC curve related to COVID-19 vs control
covid_vs_control_roc_curve <- get_roc_curve(score_df,
  control_group = "Donor control",
  positive_group = "COVID-19"
)

# ROC curve related to viral vs control
viral_vs_control_roc_curve <- get_roc_curve(score_df,
  control_group = "Donor control",
  positive_group = "Viral acute respiratory illness"
)

# ROC curve related to bacterial vs control
bact_vs_control_roc_curve <- get_roc_curve(score_df,
  control_group = "Donor control",
  positive_group = "Bacterial sepsis"
)

# plotting the ROC curves for all contrasts
p_roc <- pROC::ggroc(
  list(
    COVID = covid_vs_control_roc_curve,
    viral = viral_vs_control_roc_curve,
    bacterial = bact_vs_control_roc_curve
  ),
  size = 2
) + theme_minimal() +
  scale_colour_manual(values = c("#c9daf8ff", "#fce5cdff", "#d9ead3ff")) +
  annotate(
    "text",
    x = 0.72,
    y = 0.9,
    label = paste(
      "COVID-19 contrast \n AUC=",
      round(
        pROC::auc(covid_vs_control_roc_curve), 2
      )
    ),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.52,
    y = 0.45,
    label = paste(
      "viral respiratory contrast \n AUC=",
      round(
        pROC::auc(viral_vs_control_roc_curve), 2
      )
    ),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.4,
    y = 0.1,
    label = paste(
      "bacterial sepsis contrast \n AUC=",
      round(pROC::auc(
        bact_vs_control_roc_curve
      ), 3)
    ),
    hjust = 0
  ) +
  theme_Publication() + theme(legend.position = "none") +
  ggtitle("GSE163151 validation") +
  theme(text = element_text(size = rel(4.5))) +
  theme(plot.title = element_text(size = rel(3.5)))

p_roc
```


### Signature performance in GSE149689
This dataset includes samples from COVID-19 patients and from Flu patients. Thus, it makes it possible to simultaneously test  COVID-19 detection and cross-reactivity with flu. 

```{r, message = FALSE}
dataset <- readRDS(file = "../../data/mRNA_studies/COVID19_validation/GSE149689_pseudobulk_MI.RDS")
COVID19_signature_up_expression <- dataset$expr[COVID19_signature_up, ]
COVID19_signature_down_expression <- dataset$expr[COVID19_signature_down, ]

sample_score <- scale(
  apply(COVID19_signature_up_expression, 2, function(x) {
    geom_mean(x)
  }) -
    apply(COVID19_signature_down_expression, 2, function(x) {
      geom_mean(x)
    })
)

score_df <- data.frame(
  score = sample_score,
  group = dataset$pheno$covid_status
) %>%
  dplyr::mutate(group = relevel(group, ref = "Healthy"))


# ROC curve related to COVID-19 vs control
covid_vs_control_roc_curve <- get_roc_curve(score_df,
  control_group = "Healthy",
  positive_group = "COVID-19"
)

# ROC curve related to viral vs control
viral_vs_control_roc_curve <- get_roc_curve(score_df,
  control_group = "Healthy",
  positive_group = "Flu"
)


p_roc_GSE149689 <- pROC::ggroc(
  list(
    COVID = covid_vs_control_roc_curve,
    viral = viral_vs_control_roc_curve
  ),
  size = 2
) + theme_minimal() +
  scale_colour_manual(values = c("#c9daf8ff", "#fce5cdff")) +
  annotate(
    "text",
    x = 0.72,
    y = 0.9,
    label = paste(
      "COVID-19 contrast \n AUC=",
      round(pROC::auc(covid_vs_control_roc_curve), 2)
    ),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 0.46,
    y = 0.45,
    label = paste(
      "viral respiratory contrast \n AUC=",
      round(pROC::auc(viral_vs_control_roc_curve), 2)
    ),
    hjust = 0
  ) +
  theme_Publication() + theme(legend.position = "none") +
  ggtitle("GSE149689 validation") + theme(text = element_text(size = rel(4.5))) +
  theme(plot.title = element_text(size = rel(3.5)))
p_roc_GSE149689
```


### Global evaluation for all COVID-19 and non-COVID-19 validation studies.
```{r message = FALSE}
filter_object <-
  readRDS("../../data/mRNA_studies/filter_object.RDS")

all_contrasts <-
  readRDS(file = "../../data/mRNA_studies/all_contrasts.RDS")
all_contrasts_dict <-
  readRDS(file = "../../data/mRNA_studies/all_contrasts_dict.RDS")
levels(all_contrasts_dict$class1)[4] <- "non-infectious"

validation_contrasts <- all_contrasts[which(all_contrasts_dict$use == "validation")]
validation_contrasts_dict <- all_contrasts_dict %>% filter(use == "validation")

COVID19_signature <- list(
  up = COVID19_signature_up,
  down = COVID19_signature_down
)

COVID19_validation_contrasts_AUC <- compute_contrasts_AUCs(
  COVID19_signature,
  filter_object,
  validation_contrasts,
  validation_contrasts_dict
)


# generating summary statistics (first quartile, median, third quartile)
COVID19_validation_contrasts_AUC %>%
  group_by(class1, use) %>%
  summarise(perfomance = quantile(AUC, probs = c(0.25, 0.5, 0.75)))

p_boxplot <- get_AUC_distribution_boxplot(COVID19_validation_contrasts_AUC)

p_boxplot

compute_AUC_distribution_p_values(COVID19_validation_contrasts_AUC)
```


### Comparison with four previously published COVID-19 signatures.
```{r message = FALSE, fig.height = 15, fig.width = 8}
published_signatures = readRDS('../../data/published_signatures/published_signatures.RDS')

published_signatures = published_signatures[c(4, 1, 3, 2)]
names(published_signatures) = paste('signature', 1:4)
published_signatures_validation_AUC = lapply(published_signatures, 
                                             function(x) compute_contrasts_AUCs(x,
                                                                                filter_object, 
                                                                                validation_contrasts,
                                                                                validation_contrasts_dict))

#from Thair, exclude the discovery study
published_signatures_validation_AUC[[2]] = published_signatures_validation_AUC[[2]] %>% 
  filter(!grepl('GSE152641', study))


#plotting distributions
published_signatures_validation_AUC_long = published_signatures_validation_AUC %>%
  plyr::ldply((rbind))

p_boxplots = get_AUC_distribution_boxplot(published_signatures_validation_AUC_long) + 
  geom_point(aes(x = 1, y = -0.12), shape=22, fill="#c9daf8ff", 
             color="#c9daf8ff", size=5, stroke=3) + 
  geom_point(aes(x = 2, y = -0.12), shape=22, fill="white", 
             color="#fce5cdff", size=5, stroke=3) + 
  geom_point(aes(x = 3, y = -0.12), shape=22, 
             fill="white", color="#d9ead3ff", size=5, stroke=3) + 
  geom_point(aes(x = 4, y = -0.12), shape=22, 
             fill="white", color="#ead1dcff", size=5, stroke=3) + 
  facet_wrap(~.id, ncol = 1) + ylim(c(-0.20, 1)) + 
  scale_y_continuous(limits =c(-0.25, 1.25),
                     breaks = seq(-0.25, 1.25, by = 0.25),
                     labels = c('', seq(0, 1.0, by = 0.25), ''))


#generating summary statistics (first quartile, median, third quartile)
lapply(published_signatures_validation_AUC, function(x) 
  x %>%
    group_by(class1, use) %>%
    summarise(perfomance = quantile(AUC, probs = c(0.25, 0.5, 0.75)), .groups = 'drop')
  )

p_boxplots
lapply(published_signatures_validation_AUC, compute_AUC_distribution_p_values)
```


```{r message = FALSE}
sessionInfo()
```




